---
title: "Default Rmarkdown Template"
author: "Alix de Jersey and Jack Rivers-Auty"
date: "`r Sys.Date()`"
output:
    html_document:
      toc: true
      toc_title: "Table of Contents"
      number_sections: true
---

```{r CSS install, include=T, results='asis', comment=NA, echo=F}
x <- tryCatch(readLines("https://raw.githubusercontent.com/JackAuty/RmarkdownStyle/main/styles.css"), error=function(e) ""); if (length(x)) cat("<style>\n", paste(x, collapse="\n"), "\n</style>")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,       # Show code
  warning = FALSE,   # Suppress warnings
  message = FALSE    # Suppress messages
)

options(scipen =999 )
```

# Load Required Packages

```{r load-packages, warning=FALSE, message=FALSE, results='hide' }
# Load packages, install if required
packages <- c("ggplot2", "viridisLite", "sjPlot", "ggfortify", "readr", "MASS", "DT", "corrplot", "emmeans", "mediation", "car", "lubridate", "PupillometryR", "gghalves")
invisible(lapply(packages, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  library(pkg, character.only = TRUE)
}))
```
# Citations
```{r}

# Use the existing package list
pkg <- packages 

# Create a character vector for clean citations
citation_list <- character()  

# Loop through each package and clean citation
for (i in pkg) {
  # Get the plain text citation
  raw <- paste(capture.output(print(citation(i), style = "text", bibtex = FALSE)), collapse = "\n")
  
  # Clean: remove _, *, " and trailing . or ,.
  clean <- gsub("_", "", raw, fixed = TRUE)
  clean <- gsub("\\*", "", clean)
  clean <- gsub('\\"', "", clean)            # Handles escaped quotes
  clean <- gsub('"', "", clean)              # Handles regular quotes
  clean <- gsub("\\\\texttt\\{(.*?)\\}", "\\1", clean)
  clean <- gsub(", \\.$", "", clean)
  clean <- gsub("\\s*\\.$", "", clean)
  clean <- gsub("[\u201C\u201D]", "", clean)
  clean <- trimws(clean)

  # Append to citation list
  citation_list <- c(citation_list, clean)
}

# Display as a table
datatable(
  data.frame(References = citation_list)
)
```


# Load data
```{r}
setwd("C:/Users/ampde/OneDrive - University of Tasmania/PhD/00 Projects/2025 Blood Chem update/Blood chemistry")
blood_data <- read_csv("FFSH blood chem - 2014 to 2017 & 2025.csv")
blood_data$Date <- as.Date(blood_data$Date, format = "%d-%b-%y")
blood_data$Year <- format(as.Date(blood_data$Date), "%Y")

# Define plastic category
blood_data$Plastic <- ifelse(blood_data$Plas.mass < 0.1, "Low Plastic", "High Plastic")
```

# Glucose
```{r}
# Conversion factor
mmol_to_mgdl <- 18.01559

# Create the new glucose column
blood_data$glucose <- NA  # initialize with NA

# For rows where Glu_I has a value, convert mmol/L to mg/dL
blood_data$glucose[!is.na(blood_data$Glu_I)] <- blood_data$Glu_I[!is.na(blood_data$Glu_I)] * mmol_to_mgdl

# For rows where Glu_Z has a value, just use it directly (already in mg/dL)
blood_data$glucose[!is.na(blood_data$Glu_Z)] <- blood_data$Glu_Z[!is.na(blood_data$Glu_Z)]
```

```{r}
glucose_plot <- ggplot(blood_data, aes(x = Year, y = glucose, fill = Plastic)) +
  geom_boxplot(position = position_dodge(0.8)) +
  scale_fill_manual(values = c("High Plastic" = "#56b4e9", "Low Plastic" = "#009e73")) +
  theme_classic()
glucose_plot
```

# 2025 data
```{r}
blood_data_2025 <- subset(blood_data, Year == "2025")
```

```{r}
cols_to_test <- c(21, 17, 6:9, 12:16)
group_factor <- as.factor(blood_data_2025[[25]])

# Subset the numeric columns you want to test
numeric_data <- blood_data_2025[, cols_to_test]

# Ensure they are numeric (this should be unnecessary unless you're unsure)
numeric_data <- data.frame(lapply(numeric_data, as.numeric))


# Calculate correlation matrix
cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

# Exclude both Plas.mass (col 1) and Mass (adjust column index or name as needed)
pca_input <- numeric_data[, !(names(numeric_data) %in% c("Plas.mass", "Mass"))]

# Run PCA
pca_result <- prcomp(pca_input, scale. = TRUE)

# Extract PCA scores
pca_scores <- as.data.frame(pca_result$x)

# Add Plas.mass and Mass back in
pca_scores$Plas.mass <- numeric_data$Plas.mass
pca_scores$Mass <- numeric_data$Mass

ggplot(pca_scores, aes(x = PC1, y = PC2, color = Plas.mass, size = Mass)) +
  geom_point(alpha = 0.8) +
  scale_color_viridis_c() +
  labs(title = "PCA: Colored by Plastic Mass, Sized by Body Mass",
       x = paste0("PC1 (", round(summary(pca_result)$importance[2,1]*100, 1), "% variance)"),
       y = paste0("PC2 (", round(summary(pca_result)$importance[2,2]*100, 1), "% variance)"),
       color = "Plastic Mass",
       size = "Mass (g)") +
  theme_classic()


# Plot
corrplot::corrplot(cor_matrix, method = "circle")

custom_labels <- c(
  "Ingested plastic mass (g)", 
  "Body mass (g)", 
  "Sodium", 
  "Potassium", 
  "Chloride", 
  "Carbon dioxide", 
  "Glucose", 
  "Calcium", 
  "Anion Gap", 
  "Hematocrit",
  "Hemoglobin"
)
hist(blood_data_2025$glucose)
colnames(cor_matrix) <- custom_labels
rownames(cor_matrix) <- custom_labels

# Mixed plot: color in lower, scatterplots in upper triangle
corrplot.mixed(cor_matrix,
               lower = "circle",     # Lower triangle as color
               upper = "number",   # Upper triangle as scatterplots
               tl.col = "black",    # Text label color
               tl.pos = "lt", # Label position
               tl.labels = custom_labels, 
               diag = "n")          # No diagonal

results <- data.frame(
  Column = character(),
  Estimate = numeric(),
  StdError = numeric(),
  z_value = numeric(),
  p_value = numeric(),
  stringsAsFactors = FALSE
)

for (col_num in cols_to_test) {
  # Prepare data, remove rows with NA in either predictor or response
  dat <- data.frame(
    response = blood_data_2025[[col_num]],
    group = group_factor
  )
  dat <- na.omit(dat)
  
  if (length(unique(dat$group)) == 2) {
    # Fit GLM with Gaussian family (normal distribution)
    fit <- glm(response ~ group, data = dat, family = gaussian())
     #Diagnostic plots 
    print(autoplot(fit,which = c(1,2,4)))
    # Extract summary coefficients for the group effect (usually 2nd row)
    coef_summary <- summary(fit)$coefficients
    
    # Row 2 is group effect if factor has two levels
    est <- coef_summary[2, "Estimate"]
    se <- coef_summary[2, "Std. Error"]
    zval <- coef_summary[2, "t value"]  # For gaussian glm, it's t value
    pval <- coef_summary[2, "Pr(>|t|)"]
    
    results <- rbind(results, data.frame(
      Column = names(blood_data_2025)[col_num],
      Estimate = est,
      StdError = se,
      z_value = zval,
      p_value = pval
    ))
  }
}

print(results)
```

```{r}
ggplot(blood_data_2025, aes(x = Year, y = glucose, fill = Plastic)) +
  geom_boxplot(position = position_dodge(0.8)) +
  scale_fill_manual(values = c("High Plastic" = "#56b4e9", "Low Plastic" = "#009e73")) +
  theme_classic()
```









```{r}
# Example variables:
plastics_col <- 21  # replace with your plastics column index or name
response_col <- 12  # replace with the response variable column you want to model
colnames(blood_data_2025)
# Extract data, removing NAs
df_sub <- blood_data_2025[, c(plastics_col, response_col)]
names(df_sub) <- c("plastics", "response")
df_sub <- df_sub[complete.cases(df_sub), ]

# Add a small constant to plastics to avoid log(0)
df_sub$log_plastics <- log(df_sub$plastics + 1e-6)

# Fit linear regression with log(plastics)
fit <- lm(response ~ log_plastics, data = df_sub)

# Create prediction data for smooth line
pred_data <- data.frame(
  log_plastics = seq(min(df_sub$log_plastics), max(df_sub$log_plastics), length.out = 100)
)
pred_data$response_pred <- predict(fit, newdata = pred_data)

# Define threshold on plastics (e.g. plastics = 2)
threshold_value <- 0.5
# Add mass group
df_sub$mass_group <- ifelse(blood_data_2025$Mass < 550, "<550g", "≥550g")

# Plot with color mapped to mass_group
ggplot(df_sub, aes(x = plastics, y = response, color = mass_group)) +
  geom_point() +
  geom_line(data = transform(pred_data, plastics = exp(log_plastics)), 
            aes(x = plastics, y = response_pred), color = "#0072b2", inherit.aes = FALSE) +
  geom_vline(xintercept = threshold_value, linetype = "dashed", color = "#999999") +
  geom_smooth(method =lm)+
  scale_x_continuous(trans = "log1p") +  # log1p to match log(plastics + 1) transform
  scale_color_manual(values = c("<550g" = "#E69F00", "≥550g" = "#009E73")) +
  labs(
    title = "Regression of Response on Log Plastics",
    x = "Plastics (log scale)",
    y = "Response",
    color = "Mass Group"
  ) +
  theme_classic()

```

```{r}
ggplot(data = blood_data_2025, aes(x = Cl, y = Na, colour = Plastic)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text(aes(label = ID), vjust = -0.5, size = 3, show.legend = FALSE) +
  theme_classic()

```

data cleaning: plot the acucheck and the iSTAT as a geom_smooth (method = lm) 1/18 slope. should be tight. Do the autoplot function, it will have a high cooks distance (over 0.4). 


We might need to remove the low weight animals or later we can adjust for the low weight <550g.
To adjust for them, do a linear regression on the dependent variable (e.g., chloride) vs. weight.
Take the residuals from that and map that against plastic.mass. 
After we have adjusted for weight, what is leftover to be explained by plastic. 
The z-score axis can be hard to infer from. 
Could eventually do biomarker stuff, plot two of the strongest predictors and colour by plastic. 

How to bring in the different years? 


```{r}
colnames(blood_data_2025)

# Example variables:
plastics_col <- 21  # replace with your plastics column index or name
response_col <- 12  # replace with the response variable column you want to model

# Extract data, removing NAs
df_sub <- blood_data_2025[, c(plastics_col, response_col)]
names(df_sub) <- c("plastics", "response")

# Add a small constant to plastics to avoid log(0)
df_sub$log_plastics <- log(df_sub$plastics + 1e-6)

# Fit linear regression with log(plastics)
aov_base <- lm(response ~ log_plastics, data = df_sub)
autoplot(aov_base,which = c(1,2,4))

# BoxCox transformation 
boxcox<-boxcox(aov_base,lambda = seq(-5, 5, 1/1000), plotit = F)

Selected.lambda <-boxcox$x[boxcox$y==max(boxcox$y)] 
Selected.lambda 

# Run the transformed model 
aov_base_trans <- lm(response^Selected.lambda ~ log_plastics, data = df_sub)

autoplot(aov_base_trans, which = c(1,2,4))

# Summary of the linear model gives you the coefficients 
tab_model(aov_base_trans) 
tab_df(anova(aov_base_trans),show.rownames =T) 
```

```{r}
# Define plastics column and the response columns
plastics_col <- 21
response_cols <- setdiff(6:16, c(10, 11))  # exclude 10 and 11

# Create empty list to store results
results_list <- list()

# Loop over the response columns
for (col in response_cols) {
  
  response_name <- colnames(blood_data_2025)[col]
  
  # Extract and prepare data
  df_sub <- blood_data_2025[, c(plastics_col, col)]
  names(df_sub) <- c("plastics", "response")
  df_sub <- na.omit(df_sub)  # Remove rows with NA
  
  # Skip if not enough observations
  if (nrow(df_sub) < 3) {
    results_list[[response_name]] <- NA
    next
  }
  
  # Add small constant to plastics
  df_sub$log_plastics <- log(df_sub$plastics + 1)
  
  # Fit base model
  aov_base <- lm(response ~ log_plastics, data = df_sub)
  
  # Try BoxCox transformation
  boxcox_out <- boxcox(aov_base, lambda = seq(-5, 5, 1/1000), plotit = FALSE)
  lambda <- boxcox_out$x[which.max(boxcox_out$y)]
  
  # Fit transformed model
  if (lambda == 0) {
    response_trans <- log(df_sub$response)
  } else {
    response_trans <- df_sub$response^lambda
  }
  
  aov_base_trans <- lm(response_trans ~ log_plastics, data = df_sub)
  print(tab_model(aov_base_trans))
  aov_res <- anova(aov_base_trans)
  print(autoplot(aov_base_trans))
  # Extract p-value for log_plastics term
  p_val <- aov_res["log_plastics", "Pr(>F)"]
  
  results_list[[response_name]] <- p_val
}

# Convert to dataframe
results_df <- data.frame(
  Response = names(results_list),
  P_Value = unlist(results_list)
)

print(results_df)

```


```{r}
plot_raw_scatter_all <- function(data, plastics_col, response_cols, id_col) {
  response_cols <- setdiff(response_cols, c(10, 11))  # exclude unwanted columns
  
  plots <- list()
  
  for (col in response_cols) {
    response_name <- colnames(data)[col]
    
    df_sub <- data[, c(id_col, plastics_col, col)]
    names(df_sub) <- c("ID", "plastics", "response")
    df_sub <- na.omit(df_sub)
    
    if (nrow(df_sub) < 3) next  # skip if too few points
    
    df_sub$log_plastics <- (df_sub$plastics + 0)
    
    p <- ggplot(df_sub, aes(x = log_plastics, y = response, label = ID)) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE, color = "blue") +
      geom_text(size = 3, vjust = -0.5, check_overlap = TRUE) +
      labs(
        title = response_name,
        x = "log(Plastics)",
        y = response_name
      ) +
      theme_classic()
    
    plots[[response_name]] <- p
  }
  
  return(plots)
}

all_plots <- plot_raw_scatter_all(blood_data_2025, plastics_col = 21, response_cols = 6:16, id_col = 1)

# Show all individually
for (plot_name in names(all_plots)) {
  print(all_plots[[plot_name]])
}

```


```{r}
blood_chem_glucose <- read.csv("250721 glucose check.csv")

# Prepare the data
df_glucose <- blood_chem_glucose[, c("ID", "Gl_iSTAT", "Gl_Aqu")]
df_glucose <- na.omit(df_glucose)

# Plot Gl_iSTAT vs Gl_Aqu
ggplot(df_glucose, aes(x = Gl_iSTAT, y = Gl_Aqu, label = ID)) +
  geom_point() +
  geom_text(size = 3, vjust = -0.5, check_overlap = TRUE) +
  labs(
    title = "Comparison of Glucose Measures",
    x = "Glucose (iSTAT)",
    y = "Glucose (Aqucheck)"
  ) +
  theme_classic()
```

```{r}
plot_raw_scatter_all <- function(data, plastics_col, response_cols, id_col, mass_col, mass_threshold = 550) {
  response_cols <- setdiff(response_cols, c(10, 11))  # exclude unwanted columns
  plots <- list()
  
  for (col in response_cols) {
    response_name <- colnames(data)[col]
    
    df_sub <- data[, c(id_col, plastics_col, col, mass_col)]
    names(df_sub) <- c("ID", "plastics", "response", "mass")
    df_sub <- na.omit(df_sub)
    
    if (nrow(df_sub) < 3) next  # skip if too few points
    
    # Add mass group
    df_sub$mass_group <- ifelse(df_sub$mass < mass_threshold, "<550g", "≥550g")
    
    # Optional: transform plastics if needed (currently raw)
    df_sub$log_plastics <- (df_sub$plastics + 0)
    
    p <- ggplot(df_sub, aes(x = log_plastics, y = response, color = mass_group, label = ID)) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE) +
      geom_text(size = 3, vjust = -0.5, check_overlap = TRUE) +
      scale_color_manual(values = c("<550g" = "#E69F00", "≥550g" = "#009E73")) +
      labs(
        title = response_name,
        x = "log(Plastics)",
        y = response_name,
        color = "Mass Group"
      ) +
      theme_classic()
    
    plots[[response_name]] <- p
  }
  
  return(plots)
}


all_plots <- plot_raw_scatter_all(
  data = blood_data_2025,
  plastics_col = 21,
  response_cols = 6:16,
  id_col = 1,
  mass_col = 17  # or whatever column holds Mass
)

# Show each plot individually
for (plot_name in names(all_plots)) {
  print(all_plots[[plot_name]])
}

```

```{r}
plot_mass_by_response <- function(data, mass_col, response_cols, id_col, plastics_col = NULL, plastic_threshold = 0.5) {
  response_cols <- setdiff(response_cols, c(10, 11))
  plots <- list()
  
  for (col in response_cols) {
    response_name <- colnames(data)[col]
    
    if (!is.null(plastics_col)) {
      df_sub <- data[, c(id_col, mass_col, col, plastics_col)]
      names(df_sub) <- c("ID", "mass", "response", "plastics")
      df_sub$plastics_group <- ifelse(df_sub$plastics < plastic_threshold, "<0.5", "≥0.5")
    } else {
      df_sub <- data[, c(id_col, mass_col, col)]
      names(df_sub) <- c("ID", "mass", "response")
    }
    
    df_sub <- na.omit(df_sub)
    if (nrow(df_sub) < 3) next
    
    p <- ggplot(df_sub, aes(x = mass, y = response, label = ID)) +
      geom_point(aes(color = if (!is.null(plastics_col)) plastics_group else NULL), size=3) +
      geom_smooth(method = "lm", se = FALSE) +
      geom_text(size = 3, vjust = -0.5, check_overlap = TRUE) +
      labs(
        title = paste("Mass vs", response_name),
        x = "Mass (g)",
        y = response_name,
        color = if (!is.null(plastics_col)) "Plastics Group" else NULL
      ) +
      theme_classic()
    
    if (!is.null(plastics_col)) {
      p <- p + scale_color_manual(values = c("<0.5" = "#990099", "≥0.5" = "#0072b2"))
    }
    
    plots[[response_name]] <- p
  }
  
  return(plots)
}

mass_plots <- plot_mass_by_response(
  data = blood_data_2025,
  mass_col = 17,
  response_cols = 6:16,
  id_col = 1,
  plastics_col = 21  # optional for coloring
)

# Show each plot individually
for (plot_name in names(mass_plots)) {
  print(mass_plots[[plot_name]])
}


```

```{r}
# Define relevant columns
plastics_col <- 21
mass_col <- 17
response_cols <- setdiff(6:16, c(10, 11))  # exclude 10 and 11

# Store results
two_way_results <- data.frame(
  Response = character(),
  P_plastics = numeric(),
  P_mass = numeric(),
  P_interaction = numeric(),
  stringsAsFactors = FALSE
)

# Loop over response variables
for (col in response_cols) {
  
  response_name <- colnames(blood_data_2025)[col]
  
  # Extract required columns
  df_sub <- blood_data_2025[, c(plastics_col, mass_col, col)]
  names(df_sub) <- c("plastics", "mass", "response")
  df_sub <- na.omit(df_sub)
  
  if (nrow(df_sub) < 3) next
  
  # Log-transform plastics only
  df_sub$log_plastics <- log(df_sub$plastics + 1)
  
  # Fit initial model for BoxCox on the response
  base_model <- lm(response ~ log_plastics * mass, data = df_sub)
  boxcox_out <- boxcox(base_model, lambda = seq(-5, 5, 1/1000), plotit = FALSE)
  lambda <- boxcox_out$x[which.max(boxcox_out$y)]
  
  # Apply BoxCox transformation to the response
  df_sub$response_trans <- if (lambda == 0) log(df_sub$response) else df_sub$response^lambda
  
  # Fit two-way ANOVA on transformed response
  model <- lm(response_trans ~ log_plastics * mass, data = df_sub)
  aov_res <- anova(model)
  
  # Extract p-values
  p_plastics <- aov_res["log_plastics", "Pr(>F)"]
  p_mass <- aov_res["mass", "Pr(>F)"]
  p_interaction <- aov_res["log_plastics:mass", "Pr(>F)"]
  
  # Store results
  two_way_results <- rbind(
    two_way_results,
    data.frame(
      Response = response_name,
      P_plastics = p_plastics,
      P_mass = p_mass,
      P_interaction = p_interaction
    )
  )
}

# View the final table
print(two_way_results)
#suggests there is a starvation and non-starvation effect going on (potenitally). 
```





plot calcium for the 2014 birds with the appropriate adjustments. 


For the interaction:
Create a linear model for the response variable vs bird mass. extract the residuals from that model and plot that against the plastic mass. plas.mass not logged. 

Essentially a z score on one axis (-2, +2) and plas mass on the other. Glucose and K is the only one we should really see anything.  


```{r}
# Define relevant columns
plastics_col <- 21        # Plastic ingestion (not logged in residual plot)
mass_col <- 17            # Individual body mass
response_cols <- setdiff(6:16, c(10, 11))  # Exclude 10 and 11

# Create empty list to store results
results_list <- list()

# Loop over response variables
for (col in response_cols) {
  
  response_name <- colnames(blood_data_2025)[col]
  
  # Extract relevant data
  df_sub <- blood_data_2025[, c(plastics_col, mass_col, col)]
  names(df_sub) <- c("plastic_mass", "mass", "response")
  df_sub <- na.omit(df_sub)
  
  # Skip if not enough observations
  if (nrow(df_sub) < 3) {
    results_list[[response_name]] <- NA
    next
  }
  
  # ----------------------------------------
  # BoxCox model with plastic_mass
  df_sub$plastics <- (df_sub$plastic_mass)
  aov_base <- lm(response ~ plastics, data = df_sub)
  
  boxcox_out <- boxcox(aov_base, lambda = seq(-5, 5, 1/1000), plotit = FALSE)
  lambda <- boxcox_out$x[which.max(boxcox_out$y)]
  
  # Apply transformation
  if (lambda == 0) {
    response_trans <- log(df_sub$response)
  } else {
    response_trans <- df_sub$response^lambda
  }
  
  aov_base_trans <- lm(response_trans ~ plastics, data = df_sub)
  print(tab_model(aov_base_trans))
  print(autoplot(aov_base_trans))
  
  # Extract p-value for plastics
  aov_res <- anova(aov_base_trans)
  p_val <- aov_res["plastics", "Pr(>F)"]
  results_list[[response_name]] <- p_val
  
  # ----------------------------------------
  # Linear model: response ~ mass (body weight)
  lm_mass <- lm(response ~ mass, data = df_sub)
  df_sub$residuals_mass <- resid(lm_mass)
  
  # Plot residuals vs unlogged plastic mass
  p <- ggplot(df_sub, aes(x = plastics, y = residuals_mass)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = paste("Residuals of", response_name, "vs Plastic Mass"),
      x = "Plastic Mass (g)",
      y = "Residuals from response ~ Body Mass"
    ) +
    theme_classic()
  
  print(p)


  p <- ggplot(df_sub, aes(x = plastics, y = response)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "#0072b2") +
    labs(
      title = paste("Raw data from ", response_name, "vs Plastic Mass"),
      x = "Plastic Mass (g)",
      y = "Response"
    ) +
    theme_classic()
  
  print(p)
  
  file_name <- paste0("plot_", gsub(" ", "_", response_name), ".pdf")

  # Save as PDF
  ggsave(
    filename = file_name,
    plot = p,
    device = "pdf",
    width = 6,
    height = 5
  )
}

# Convert p-values to dataframe
results_df <- data.frame(
  Response = names(results_list),
  P_Value = unlist(results_list)
)

print(results_df)

# Adjust p-values
results_df$P_FDR <- as.numeric(p.adjust(results_df$P_Value, method = "fdr"))
# Print final results
print(results_df)
datatable(results_df) %>%
  formatRound(columns = c("P_Value", "P_FDR"), digits = 3)

```

```{r}
# Define relevant columns
plastics_col <- 21        # Plastic ingestion (unlogged)
mass_col <- 17            # Body mass
response_cols <- setdiff(6:16, c(10, 11))  # Response variables

# Loop through each response variable
for (col in response_cols) {
  
  response_name <- colnames(blood_data_2025)[col]
  
  # Subset and rename data
  df_sub <- blood_data_2025[, c(plastics_col, mass_col, col)]
  names(df_sub) <- c("plastic_mass", "mass", "response")
  df_sub <- na.omit(df_sub)
  
  if (nrow(df_sub) < 3) next
  
  # Fit adjusted model: response ~ plastic_mass + mass
  adjusted_model <- lm(response ~ plastic_mass * mass, data = df_sub)
  
  # Print model summary
  cat("\n--- Adjusted model for:", response_name, "---\n")
  print(summary(adjusted_model))
  
  # Optional plot of partial effect of plastic_mass
  p <- ggplot(df_sub, aes(x = plastic_mass, y = response)) +
    geom_point() +
    geom_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "darkgreen") +
    labs(
      title = paste("Adjusted:", response_name, "~ Plastic Mass + Mass"),
      x = "Plastic Mass (g)",
      y = response_name
    ) +
    theme_classic()
  
  print(p)
}

```


```{r}
# Define relevant columns
plastics_col <- 21        # Plastic ingestion (unlogged)
mass_col <- 17            # Body mass
response_cols <- setdiff(6:16, c(10, 11))# Response variables
ID <- 1

# Loop through each response variable
for (col in response_cols) {
  
  response_name <- colnames(blood_data_2025)[col]
  
  # Subset and rename data
  df_sub <- blood_data_2025[, c(plastics_col, mass_col, col, ID)]
  names(df_sub) <- c("plastic_mass", "mass", "response", "ID")
  df_sub <- na.omit(df_sub)
  
  if (nrow(df_sub) < 3) next
  
  # Step 1: Fit response ~ mass
  model_mass <- lm(response ~ mass, data = df_sub)
  
  # Step 2: Extract residuals
  df_sub$resid_mass <- resid(model_mass)
  
  # Step 3: Plot residuals vs unlogged plastic mass
  p <- ggplot(df_sub, aes(x = plastic_mass, y = resid_mass, label = ID)) +
        geom_point() +
        geom_text(vjust = -0.5, size = 3) +  # add labels just above points
        geom_smooth(method = "lm", se = TRUE, color = "blue") +
        labs(
          title = paste("Residuals of", response_name, "~ Mass vs Plastic Mass"),
          x = "Plastic Mass (g)",
          y = "Residuals from response ~ Mass"
        ) +
        theme_classic()
      
        
  print(p)
}

```


```{r}
# Define relevant columns
plastics_col <- 21        # Plastic ingestion
mass_col <- 17            # Body mass
response_cols <- setdiff(6:16, c(10, 11))  # Response variables

# Create empty list to store results
results_list <- list()

for (col in response_cols) {
  
  response_name <- colnames(blood_data_2025)[col]
  
  # Subset and rename relevant data
  df_sub <- blood_data_2025[, c(plastics_col, mass_col, col)]
  names(df_sub) <- c("plastic_mass", "mass", "response")
  df_sub <- na.omit(df_sub)
  
  if (nrow(df_sub) < 5) {
    results_list[[response_name]] <- NA
    next
  }
  
  # Step 1: Model 1 - response ~ mass
  model_mass <- lm(response ~ mass, data = df_sub)
  df_sub$resid_mass <- resid(model_mass)

  # Step 2: Model 2 - residuals ~ plastic_mass
  model_resid_plastic <- lm(resid_mass ~ plastic_mass, data = df_sub)
  print(paste("Residuals ~ Plastic Mass for", response_name))
  print(summary(model_resid_plastic))
  
  # Step 3: Adjusted Model - response ~ mass * plastic_mass
  model_adjusted <- lm(response ~ mass * plastic_mass, data = df_sub)
  print(paste("Adjusted model for", response_name))
  print(summary(model_adjusted))
  
  # Store p-value of interaction term
  interaction_p <- anova(model_adjusted)["mass:plastic_mass", "Pr(>F)"]
  results_list[[response_name]] <- interaction_p
}

# Compile results
results_df <- data.frame(
  Response = names(results_list),
  Interaction_P = unlist(results_list)
)
results_df$P_FDR <- p.adjust(results_df$Interaction_P, method = "fdr")
print(results_df)

```

```{r}
par(mfrow=c(1,1))
# Extract relevant data (assuming cols 1 = Plas.mass, 2 = Mass, 3:11 = response variables)
numeric_data <- blood_data_2025[, c(21, 17, 6:9, 12:16)]
numeric_data <- data.frame(lapply(numeric_data, as.numeric))

# Rename for clarity
colnames(numeric_data)[1:2] <- c("Plas.mass", "Mass")

# Extract Mass vector
mass_vector <- numeric_data$Mass

# Create residual matrix for columns 3:11 (excluding Plas.mass and Mass)
residual_matrix <- numeric_data[, 3:11]

# Adjust each column for mass
for (var_name in colnames(residual_matrix)) {
  df_temp <- data.frame(y = numeric_data[[var_name]], mass = mass_vector)
  df_temp <- na.omit(df_temp)

  if (nrow(df_temp) >= 5) {
    model <- lm(y ~ 1, data = df_temp)
    residuals_adj <- rep(NA, nrow(numeric_data))
    residuals_adj[as.numeric(rownames(df_temp))] <- resid(model)
    residual_matrix[[var_name]] <- residuals_adj
  } else {
    residual_matrix[[var_name]] <- NA
  }
}

# Step 5: Add Plas.mass and Mass back in (unadjusted)
residual_matrix$Plas.mass <- log(numeric_data$Plas.mass+1)
residual_matrix$Mass <- numeric_data$Mass

# Step 6: Reorder columns: Plas.mass, Mass, then adjusted variables
residual_matrix <- residual_matrix[, c("Plas.mass", "Mass", colnames(residual_matrix)[!colnames(residual_matrix) %in% c("Plas.mass", "Mass")])]

# Step 7: Calculate correlation matrix
adj_cor_matrix <- cor(residual_matrix, use = "pairwise.complete.obs")

custom_labels <- c(
  "Ingested plastic mass (g)", 
  "Body mass (g)", 
  "Sodium", 
  "Potassium", 
  "Chloride", 
  "Carbon dioxide", 
  "Glucose", 
  "Calcium", 
  "Anion Gap", 
  "Hematocrit",
  "Hemoglobin"
)

colnames(adj_cor_matrix) <- custom_labels
rownames(adj_cor_matrix) <- custom_labels

# Step 8: Plot
corrplot.mixed(adj_cor_matrix,
               lower = "circle",
               upper = "number",
               tl.col = "black",
               tl.pos = "lt",
               tl.cex = 0.9,
               tl.labels = custom_labels,
               diag = "n",
               title = "Correlation Matrix (Responses Adjusted for Mass)")
```


```{r}
library(corrplot)

par(mfrow = c(1,1))

# Extract relevant data (assuming cols 21 = Plas.mass, 17 = Mass, 6:9 & 12:16 = blood chemistry)
numeric_data <- blood_data_2025[, c(21, 17, 6:9, 12:16)]
numeric_data <- data.frame(lapply(numeric_data, as.numeric))

# Rename for clarity
colnames(numeric_data)[1:2] <- c("Plas.mass", "Mass")

# Step 1: Log-transform plastic mass if desired
numeric_data$Plas.mass <- log(numeric_data$Plas.mass + 1)

# Step 2: Calculate correlation matrix (no adjustment)
cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

# Step 3: Custom labels
custom_labels <- c(
  "Ingested plastic mass (g)", 
  "Body mass (g)", 
  "Sodium", 
  "Potassium", 
  "Chloride", 
  "Carbon dioxide", 
  "Glucose", 
  "Calcium", 
  "Anion Gap", 
  "Hematocrit",
  "Hemoglobin"
)

colnames(cor_matrix) <- custom_labels
rownames(cor_matrix) <- custom_labels

# Step 4: Plot
# Plot correlation matrix
corrplot.mixed(cor_matrix,
               lower = "circle",
               upper = "number",
               tl.col = "black",
               tl.pos = "lt",
               tl.cex = 0.9,
               diag = "n",
               title = "Correlation Matrix (Raw Data)")
```












#############################################################################################

Mediation analysis

Run a mediation analysis on the 2025 data
```{r}
cal_2025 <- data.frame(
  ID = blood_data_2025$ID,
  calcium = blood_data_2025$Ca_Z, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

initial_model <- lm(calcium ~ mass + plastic, data = cal_2025)
autoplot(initial_model)
boxcox_result <- boxcox(initial_model, lambda = seq(-10, 10, 1/300))
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
if(lambda == 0){
  cal_2025$calcium_trans <- log(cal_2025$calcium)
} else {
  cal_2025$calcium_trans <- (cal_2025$calcium^lambda)
}
hist(cal_2025$calcium)
hist(cal_2025$calcium_trans)

cal_model <- lm(calcium_trans ~ mass, data = cal_2025)

autoplot(cal_model)
cal_2025$residuals <- stdres(cal_model)

cal_2025$mass_centred <- as.numeric(scale(cal_2025$mass, center = TRUE, scale = TRUE))
cal_2025$plastic_centred <- as.numeric(scale(cal_2025$plastic, center = TRUE, scale = TRUE))


fit.direct.model.centred <- lm(calcium_trans ~ mass_centred, data = cal_2025)
fit.mediator.variable.model.centred <- lm(calcium_trans ~ mass_centred + plastic_centred, data = cal_2025)

par(mfrow=c(2,2)) 
plot(fit.direct.model)
plot(fit.mediator.variable.model)

mediator_model <- lm(plastic_centred ~ mass_centred, data = cal_2025)

mediation_result <- mediate(
  model.m = mediator_model,
  model.y = fit.mediator.variable.model.centred,
  treat = "mass_centred",
  mediator = "plastic_centred",
  boot = TRUE,
  sims = 1000
)

summary(mediation_result)
``` 

mediation with jack

```{r calcium}
cal_2025 <- data.frame(
  ID = blood_data_2025$ID,
  calcium = blood_data_2025$Ca_Z, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

cal_2025$plastic_scale<-as.numeric(scale(log(cal_2025$plastic+1)))

cal_2025$mass_scale<-as.numeric(scale(cal_2025$mass^2))
hist(cal_2025$mass_scale)


lm_diag <- lm(calcium ~plastic_scale + mass_scale, data = cal_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
if(lambda == 0){
  cal_2025$calcium_trans <- log(cal_2025$calcium)
} else {
  cal_2025$calcium_trans <- (cal_2025$calcium^lambda)
}
hist(cal_2025$calcium)
hist(cal_2025$calcium_trans)

cal_2025$calcium_trans<-scale(cal_2025$calcium_trans)

fit.direct.model.centred<-lm(calcium_trans~plastic_scale, data=cal_2025)
fit.mediator.variable.model.centred<-lm(calcium_trans~plastic_scale + mass_scale, data=cal_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
calcium_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary_calcium<- summary(calcium_mediation)
```

```{r sodium}
sodium_2025 <- data.frame(
  ID = blood_data_2025$ID,
  sodium = blood_data_2025$Na, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

sodium_2025$plastic_scale<-as.numeric(scale(log(sodium_2025$plastic+1)))
sodium_2025$mass_scale<-as.numeric(scale(sodium_2025$mass^2))
hist(sodium_2025$mass_scale)

lm_diag <- lm(sodium ~ plastic_scale + mass_scale, data = sodium_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  sodium_2025$sodium_trans <- log(sodium_2025$sodium)
} else {
  sodium_2025$sodium_trans <- (sodium_2025$sodium^lambda)
}

hist(sodium_2025$sodium)
hist(sodium_2025$sodium_trans)

sodium_2025$sodium_trans<-scale(sodium_2025$sodium_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = sodium_2025)
fit.mediator.variable.model.centred<-lm(sodium_trans~plastic_scale + mass_scale, data = sodium_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
sodium_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(sodium_mediation)
```

```{r potassium}
potassium_2025 <- data.frame(
  ID = blood_data_2025$ID,
  potassium = blood_data_2025$K, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

potassium_2025$plastic_scale<-as.numeric(scale(log(potassium_2025$plastic+1)))
potassium_2025$mass_scale<-as.numeric(scale(potassium_2025$mass^2))
hist(potassium_2025$mass_scale)


lm_diag <- lm(potassium ~ plastic_scale + mass_scale, data = potassium_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  potassium_2025$potassium_trans <- log(potassium_2025$potassium)
} else {
  potassium_2025$potassium_trans <- (potassium_2025$potassium^lambda)
}

hist(potassium_2025$potassium)
hist(potassium_2025$potassium_trans)

potassium_2025$potassium_trans<-scale(potassium_2025$potassium_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = potassium_2025)
fit.mediator.variable.model.centred<-lm(potassium_trans~plastic_scale + mass_scale, data = potassium_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
potassium_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(potassium_mediation)
```

```{r choride}
chloride_2025 <- data.frame(
  ID = blood_data_2025$ID,
  chloride = blood_data_2025$Cl, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

chloride_2025$plastic_scale<-as.numeric(scale(log(chloride_2025$plastic+1)))
chloride_2025$mass_scale<-as.numeric(scale(chloride_2025$mass^2))
hist(chloride_2025$mass_scale)

lm_diag <- lm(chloride ~ plastic_scale + mass_scale, data = chloride_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  chloride_2025$chloride_trans <- log(chloride_2025$chloride)
} else {
  chloride_2025$chloride_trans <- (chloride_2025$chloride^lambda)
}

hist(chloride_2025$chloride, breaks=12)
hist(chloride_2025$chloride_trans, breaks=12)
chloride_2025$chloride_trans<-scale(chloride_2025$chloride_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = chloride_2025)
fit.mediator.variable.model.centred<-lm(chloride_trans~plastic_scale + mass_scale, data = chloride_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
chloride_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(chloride_mediation)
```

```{r CO2}
co2_2025 <- data.frame(
  ID = blood_data_2025$ID,
  co2 = blood_data_2025$CO2, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

co2_2025$plastic_scale<-as.numeric(scale(log(co2_2025$plastic+1)))
co2_2025$mass_scale<-as.numeric(scale(co2_2025$mass^2))
hist(co2_2025$mass_scale)

lm_diag <- lm(co2 ~ plastic_scale + mass_scale, data = co2_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  co2_2025$co2_trans <- log(co2_2025$co2)
} else {
  co2_2025$co2_trans <- (co2_2025$co2^lambda)
}

hist(co2_2025$co2)
hist(co2_2025$co2_trans)

co2_2025$co2_trans<-scale(co2_2025$co2_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = co2_2025)
fit.mediator.variable.model.centred<-lm(co2_trans~plastic_scale + mass_scale, data = co2_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
co2_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(co2_mediation)
```

```{r glucose}
glucose_2025 <- data.frame(
  ID = blood_data_2025$ID,
  glucose = blood_data_2025$Glu_Z, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

glucose_2025$plastic_scale<-as.numeric(scale(log(glucose_2025$plastic+1)))
glucose_2025$mass_scale<-as.numeric(scale(glucose_2025$mass^2))
hist(glucose_2025$mass_scale)

lm_diag <- lm(glucose ~ plastic_scale + mass_scale, data = glucose_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  glucose_2025$glucose_trans <- log(glucose_2025$glucose)
} else {
  glucose_2025$glucose_trans <- (glucose_2025$glucose^lambda)
}

hist(glucose_2025$glucose)
hist(glucose_2025$glucose_trans)

glucose_2025$glucose_trans<-scale(glucose_2025$glucose_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = glucose_2025)
fit.mediator.variable.model.centred<-lm(glucose_trans~plastic_scale + mass_scale, data = glucose_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
glucose_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(glucose_mediation)
```

```{r Anion Gap}
an_2025 <- data.frame(
  ID = blood_data_2025$ID,
  an = blood_data_2025$An, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

an_2025$plastic_scale<-as.numeric(scale(log(an_2025$plastic+1)))
an_2025$mass_scale<-as.numeric(scale(an_2025$mass^2))
hist(an_2025$mass_scale)

lm_diag <- lm(an ~ plastic_scale + mass_scale, data = an_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  an_2025$an_trans <- log(an_2025$an)
} else {
  an_2025$an_trans <- (an_2025$an^lambda)
}

hist(an_2025$an)
hist(an_2025$an_trans)

an_2025$an_trans<-scale(an_2025$an_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = an_2025)
fit.mediator.variable.model.centred<-lm(an_trans~plastic_scale + mass_scale, data = an_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
an_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(an_mediation)
```

```{r hemoglobin}
hemoglobin_2025 <- data.frame(
  ID = blood_data_2025$ID,
  hemoglobin = blood_data_2025$Hb, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

hemoglobin_2025$plastic_scale<-as.numeric(scale(log(hemoglobin_2025$plastic+1)))
hemoglobin_2025$mass_scale<-as.numeric(scale(hemoglobin_2025$mass^2))
hist(hemoglobin_2025$mass_scale)

lm_diag <- lm(hemoglobin ~ plastic_scale + mass_scale, data = hemoglobin_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  hemoglobin_2025$hemoglobin_trans <- log(hemoglobin_2025$hemoglobin)
} else {
  hemoglobin_2025$hemoglobin_trans <- (hemoglobin_2025$hemoglobin^lambda)
}

hist(hemoglobin_2025$hemoglobin)
hist(hemoglobin_2025$hemoglobin_trans)

hemoglobin_2025$hemoglobin_trans<-scale(hemoglobin_2025$hemoglobin_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = hemoglobin_2025)
fit.mediator.variable.model.centred<-lm(hemoglobin_trans~plastic_scale + mass_scale, data = hemoglobin_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
hemoglobin_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(hemoglobin_mediation)
```

```{r hematocrit}
hematocrit_2025 <- data.frame(
  ID = blood_data_2025$ID,
  hematocrit = blood_data_2025$Glu_Z, 
  plastic = blood_data_2025$Plas.mass, 
  mass = blood_data_2025$Mass
)

hematocrit_2025$plastic_scale<-as.numeric(scale(log(hematocrit_2025$plastic+1)))
hematocrit_2025$mass_scale<-as.numeric(scale(hematocrit_2025$mass^2))
hist(hematocrit_2025$mass_scale)

lm_diag <- lm(hematocrit ~ plastic_scale + mass_scale, data = hematocrit_2025)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))

lambda <- boxcox_result$x[which.max(boxcox_result$y)]

if(lambda == 0){
  hematocrit_2025$hematocrit_trans <- log(hematocrit_2025$hematocrit)
} else {
  hematocrit_2025$hematocrit_trans <- (hematocrit_2025$hematocrit^lambda)
}

hist(hematocrit_2025$hematocrit)
hist(hematocrit_2025$hematocrit_trans)

hematocrit_2025$hematocrit_trans<-scale(hematocrit_2025$hematocrit_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = hematocrit_2025)
fit.mediator.variable.model.centred<-lm(hematocrit_trans~plastic_scale + mass_scale, data = hematocrit_2025)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
hematocrit_mediation<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(hematocrit_mediation)
```

Visualisation and table of 2025 mediation analyses
```{r}
# Vector of mediator names (prefixes)
mediators <- c("calcium", "chloride", "co2", "glucose", "hematocrit", "hemoglobin", "an", "sodium")

# Initialize list to store data frames
all_meds <- list()

# Loop over each mediator
for (med in mediators) {
  
  # Dynamically get the mediation object
  mediation_obj <- get(paste0(med, "_mediation"))
  
  # Get summary
  summary_med <- summary(mediation_obj)
  
  # Build and format data frame
  df_med <- data.frame(
    Mediator = med,
    Effect = c("ACME (indirect effect)",
               "ADE (direct effect)",
               "Total Effect",
               "Proportion Mediated"),
    Estimate = round(c(summary_med$d0, summary_med$z0, summary_med$tau.coef, summary_med$n0), 2),
    CI = paste0("[",
                round(c(summary_med$d0.ci[1], summary_med$z0.ci[1], summary_med$tau.ci[1], summary_med$n0.ci[1]), 3),
                ", ",
                round(c(summary_med$d0.ci[2], summary_med$z0.ci[2], summary_med$tau.ci[2], summary_med$n0.ci[2]), 3),
                "]"),
    p_value = c(summary_med$d0.p, summary_med$z0.p, summary_med$tau.p, summary_med$n0.p)
  )
  
  # Format p-values
  df_med$p_value <- ifelse(df_med$p_value < 0.0001,
                           "<0.0001",
                           sprintf("%.4f", df_med$p_value))
  
  # Append to list
  all_meds[[med]] <- df_med
}

# Combine all into one data frame
combined_df <- do.call(rbind, all_meds)

# Display interactive table
datatable(combined_df,
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: left; font-weight: bold;',
            "Combined Mediation Analysis Results"
          ),
          rownames = FALSE,
          options = list(pageLength = 20))

# Export to CSV
# write.csv(combined_df, file = "combined_mediation_results.csv", row.names = FALSE)

```

```{r}
p_stars <- function(p) ifelse(p < 0.001, "***",
                       ifelse(p < 0.01, "**",
                       ifelse(p < 0.05, "*", "")))

plot_mediation_triangle <- function(
  med_fit, out_fit,
  treat, mediator,
  x_label = treat, m_label = mediator, y_label = "Outcome",

  # colours & style
  box_fills = c(X = "#E6F0FA", M = "#FDE6E6", Y = "#F1E6FA"),
  edge_col  = "#222222",
  text_col  = "#222222",
  line_width = 1,
  arrow_head_size = 0.25,    # size of arrowhead only
  line_inset = 1.8,          # pull back arrow line from box centre

  # nudges
  a_nudge = c(-1.6, 0.8),
  b_nudge = c( 1.8, 0.8),
  c_nudge = c( 0.0,-0.7),

  digits_b = 3, digits_t = 2
) {
  sm_m <- summary(med_fit)$coefficients
  sm_y <- summary(out_fit)$coefficients

  a_est <- sm_m[treat, "Estimate"]; a_t <- sm_m[treat, "t value"]; a_df <- df.residual(med_fit); a_p <- 2*pt(-abs(a_t), a_df)
  b_est <- sm_y[mediator, "Estimate"]; b_t <- sm_y[mediator, "t value"]; b_df <- df.residual(out_fit); b_p <- 2*pt(-abs(b_t), b_df)
  c_est <- sm_y[treat,  "Estimate"]; c_t <- sm_y[treat,  "t value"]; c_df <- df.residual(out_fit); c_p <- 2*pt(-abs(c_t), c_df)

  a_lab <- sprintf("(a)\nb = %.*f, t(%d) = %.*f%s", 
                   digits_b, a_est, a_df, digits_t, a_t, p_stars(a_p))
  b_lab <- sprintf("(b)\nb = %.*f, t(%d) = %.*f%s", 
                   digits_b, b_est, b_df, digits_t, b_t, p_stars(b_p))
  c_lab <- sprintf("(c')\nb = %.*f, t(%d) = %.*f%s", 
                   digits_b, c_est, c_df, digits_t, c_t, p_stars(c_p))

  nodes <- data.frame(
    name = c(x_label, m_label, y_label),
    id   = c("X","M","Y"),
    x    = c(0, 5, 10),
    y    = c(0, 4, 0)
  )

  # shorten segments more so arrowheads are clear of boxes
  a_x  <- c(nodes$x[1] + line_inset*0.35, nodes$x[2] - line_inset*0.35)
  a_y  <- c(nodes$y[1] + line_inset*0.28, nodes$y[2] - line_inset*0.28)
  b_x  <- c(nodes$x[2] + line_inset*0.35, nodes$x[3] - line_inset*0.35)
  b_y  <- c(nodes$y[2] - line_inset*0.28, nodes$y[3] + line_inset*0.28)
  c_x  <- c(nodes$x[1] + line_inset, nodes$x[3] - line_inset)
  c_y  <- c(0, 0)

  # label midpoints + nudges
  mid_a <- c(mean(a_x) + a_nudge[1], mean(a_y) + a_nudge[2])
  mid_b <- c(mean(b_x) + b_nudge[1], mean(b_y) + b_nudge[2])
  mid_c <- c(mean(c_x) + c_nudge[1], mean(c_y) + c_nudge[2])

  arrow_style <- arrow(type = "closed", 
                       length = unit(arrow_head_size, "cm"))

  ggplot() +
    # a path
    geom_segment(aes(x = a_x[1], y = a_y[1], 
                     xend = a_x[2], yend = a_y[2]),
                 arrow = arrow_style, 
                 linewidth = line_width, colour = edge_col) +
    annotate("text", x = mid_a[1], y = mid_a[2], label = a_lab,
             size = 3.8, hjust = 0.5, colour = text_col) +

    # b path
    geom_segment(aes(x = b_x[1], y = b_y[1], 
                     xend = b_x[2], yend = b_y[2]),
                 arrow = arrow_style, 
                 linewidth = line_width, colour = edge_col) +
    annotate("text", x = mid_b[1], y = mid_b[2], label = b_lab,
             size = 3.8, hjust = 0.5, colour = text_col) +

    # c′ path
    geom_segment(aes(x = c_x[1], y = c_y[1], 
                     xend = c_x[2], yend = c_y[2]),
                 arrow = arrow_style, 
                 linewidth = line_width, colour = edge_col) +
    annotate("text", x = mid_c[1], y = mid_c[2], label = c_lab,
             size = 3.8, hjust = 0.5, colour = text_col) +

    # nodes
    geom_label(
      data = nodes,
      aes(x = x, y = y, label = name, fill = id),
      size = 5,
      colour = "black",
      label.size = 0.6,
      label.r = unit(0.55, "lines"),           # corner roundness
      label.padding = unit(0.5, "lines")       # inner text padding
    ) +
    scale_fill_manual(values = setNames(box_fills, c("X","M","Y")), guide = "none") +
    coord_fixed() + theme_void() +
    xlim(-1, 11) + ylim(-1.6, 5.4)
}
```

```{r}
plot_mediation_triangle(
  med_fit = fit.direct.model.centred,
  out_fit = fit.mediator.variable.model.centred,
  treat   = "plastic_scale",
  mediator= "mass_scale",
  x_label = "plastic",
  m_label = "mass",
  y_label = "bloodchem"
)
```



###############################################################################################


blood chem data 2014

```{r}
# Convert to Date class if not already
blood_data$Date <- as.Date(blood_data$Date)
blood_data$Year <- year(blood_data$Date)

# Build cleaned dataframe
blood_data_2014 <- data.frame(
  ID = blood_data$ID,
  Calcium = blood_data$Ca_I,
  Glucose = blood_data$Glu_I,
  Creatine = blood_data$Cr_I,
  Plastic = blood_data$Plas.mass,
  Mass = blood_data$Mass, 
  Year = blood_data$Year
)

blood_data_2014 <- na.omit(blood_data_2014)

plastics_col <- 5
mass_col <- 6
year_col <- 7
response_cols <- 2:4

# Create empty list to store results
results_list <- list()

for (col in response_cols) {
  
  response_name <- colnames(blood_data_2014)[col]
  
  # Subset data
  df_sub <- blood_data_2014[, c(plastics_col, mass_col, year_col, col)]
  names(df_sub) <- c("plastic_mass", "mass", "year", "response")
  
  # Step 1: Fit model for Box-Cox
  lm_prebox <- lm(response ~ mass + factor(year), data = df_sub)
  
  # Step 2: Box-Cox transformation to determine best lambda
  bc <- boxcox(lm_prebox, lambda = seq(-10, 10, 1/300), plotit = FALSE)

  lambda <- bc$x[which.max(bc$y)]
  
  # Step 3: Apply Box-Cox transformation to response
  if (lambda == 0) {
    df_sub$response_bc <- log(df_sub$response)
  } else {
    df_sub$response_bc <- (df_sub$response^lambda) / lambda
  }
  
  hist(df_sub$response)
  hist(df_sub$response_bc)
  
  # Step 4: Adjust for mass and year with transformed response
  model_adjust <- lm(response_bc ~ mass + factor(year), data = df_sub)
  df_sub$resid_adjusted <- resid(model_adjust)
  
  # Step 5: Plot residuals vs plastic mass
  p <- ggplot(df_sub, aes(x = plastic_mass, y = resid_adjusted)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = paste("Residuals of", response_name, "(Box-Cox λ =", round(lambda, 2), ") vs Plastic Mass"),
      x = "Plastic Mass (g)",
      y = "Adjusted Residuals"
    ) +
    theme_classic()
    
  print(p)
  
  # Step 6: Model plastic effect on residuals
  model_plastic <- lm(resid_adjusted ~ plastic_mass, data = df_sub)
  print(paste("Plastic effect on", response_name, "after Box-Cox + adjustment for mass and year"))
  print(summary(model_plastic))
  
  # Store p-value
  results_list[[response_name]] <- summary(model_plastic)$coefficients["plastic_mass", "Pr(>|t|)"]
}

# Compile and adjust p-values
results_df <- data.frame(
  Response = names(results_list),
  P_value = unlist(results_list)
)
results_df$P_FDR <- p.adjust(results_df$P_value, method = "fdr")

print(results_df)
```

```{r calcium 2014 mediation analysis}
cal_2014 <- data.frame(
  ID = blood_data$ID,
  calcium = blood_data$Ca_I, 
  plastic = blood_data$Plas.mass, 
  mass = blood_data$Mass
)

cal_2014 <- na.omit(cal_2014)

cal_2014$plastic_scale<-as.numeric(scale(log(cal_2014$plastic+1)))

cal_2014$mass_scale<-as.numeric(scale(cal_2014$mass^2))
hist(cal_2014$mass_scale)
hist(cal_2014$plastic_scale)

lm_diag <- lm(calcium ~plastic_scale + mass_scale, data = cal_2014)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
if(lambda == 0){
  cal_2014$calcium_trans <- log(cal_2014$calcium)
} else {
  cal_2014$calcium_trans <- (cal_2014$calcium^lambda)
}
hist(cal_2014$calcium)
hist(cal_2014$calcium_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = cal_2014)
fit.mediator.variable.model.centred<-lm(calcium_trans~plastic_scale + mass_scale, data = cal_2014)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
calcium_mediation_2014<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(calcium_mediation_2014)
```

```{r glucose 2014 mediation analysis}
glucose_2014 <- data.frame(
  ID = blood_data$ID,
  glucose = blood_data$Glu_I, 
  plastic = blood_data$Plas.mass, 
  mass = blood_data$Mass, 
  year = blood_data$Year
)

glucose_2014 <- na.omit(glucose_2014)

glucose_2014$plastic_scale<-as.numeric(scale(log(glucose_2014$plastic+1)))

glucose_2014$mass_scale<-as.numeric(scale(glucose_2014$mass^2))
hist(glucose_2014$mass_scale)
hist(glucose_2014$plastic_scale)

lm_diag <- lm(glucose ~plastic_scale + mass_scale, data = glucose_2014)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
if(lambda == 0){
  glucose_2014$glucose_trans <- log(glucose_2014$glucose)
} else {
  glucose_2014$glucose_trans <- (glucose_2014$glucose^lambda)
}
hist(glucose_2014$glucose)
hist(glucose_2014$glucose_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = glucose_2014)
fit.mediator.variable.model.centred<-lm(glucose_trans~plastic_scale + mass_scale, data = glucose_2014)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
glucose_mediation_2014<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(glucose_mediation_2014)
```

```{r creatine 2014 mediation analysis}
creatine_2014 <- data.frame(
  ID = blood_data$ID,
  creatine = blood_data$Cr_I, 
  plastic = blood_data$Plas.mass, 
  mass = blood_data$Mass
)

creatine_2014 <- na.omit(creatine_2014)

creatine_2014$plastic_scale<-as.numeric(scale(log(creatine_2014$plastic+1)))

creatine_2014$mass_scale<-as.numeric(scale(creatine_2014$mass^2))
hist(creatine_2014$mass_scale)
hist(creatine_2014$plastic_scale)

lm_diag <- lm(creatine ~plastic_scale + mass_scale, data = creatine_2014)

boxcox_result <- boxcox(lm_diag, lambda = seq(-10, 10, 1/300))
lambda <- boxcox_result$x[which.max(boxcox_result$y)]
if(lambda == 0){
  creatine_2014$creatine_trans <- log(creatine_2014$creatine)
} else {
  creatine_2014$creatine_trans <- (creatine_2014$creatine^lambda)
}
hist(creatine_2014$creatine)
hist(creatine_2014$creatine_trans)

fit.direct.model.centred<-lm(mass_scale~plastic_scale, data = creatine_2014)
fit.mediator.variable.model.centred<-lm(creatine_trans~plastic_scale + mass_scale, data = creatine_2014)

summary(fit.direct.model.centred)
summary(fit.mediator.variable.model.centred)

autoplot(fit.direct.model.centred)
autoplot(fit.mediator.variable.model.centred)

set.seed(333)
creatine_mediation_2014<-mediate(fit.direct.model.centred, fit.mediator.variable.model.centred, treat="plastic_scale", mediator = "mass_scale", boot=T, sims=1000)
summary(creatine_mediation_2014)
```




Figures 

```{r}
raincloud_glucose <- ggplot(glucose_allyears, aes(x = year, y = glucose, fill = group, color = group)) +
  geom_half_violin(side = "l", alpha = 0.8, width = 0.6) +
  geom_point(position = position_jitter(width = 0.1), size = 2, alpha = 0.8) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 1, color = "black", position = position_nudge(x = 0.15)) +
  theme_classic() +
  labs(
    title = "Glucose by Year",
    x = "Year",
    y = "Glucose"
  ) +
  theme(
    text = element_text(size = 14),
    legend.position = "none"
  )
raincloud_glucose

```

```{r}
# Ensure factors are in correct order
glucose_allyears$group <- factor(glucose_allyears$group, levels = c("Low Plastic", "High Plastic"))

# Make composite x-axis
glucose_allyears$year_group <- interaction(glucose_allyears$year, glucose_allyears$group, sep = "_")

raincloud_glucose <- ggplot(glucose_allyears, aes(x = year_group, y = glucose, fill = group, color = group)) +
  geom_half_violin(side = "l", alpha = 0.8, width = 0.6) +
  geom_point(position = position_jitter(width = 0.1), size = 2, alpha = 0.8) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 1, color = "black", position = position_nudge(x = 0.15)) +
  theme_classic() +
  labs(
    title = "Glucose by Year",
    x = "Year",
    y = "Glucose"
  ) +
  scale_x_discrete(labels = function(x) sub("_.*", "", x)) +  # Show just the year on x-axis
  theme(
    text = element_text(size = 14),
    legend.position = "right"
  )

raincloud_glucose

```

```{r}
ggplot(glucose_allyears, aes(x = factor(year), y = glucose, fill = group)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  scale_fill_manual(values = c("Low Plastic" = "#56b4e9", "High Plastic" = "#009e73")) +
  theme_classic() +
  labs(
    title = "Glucose by Year and Group",
    x = "Year",
    y = "Glucose",
    fill = "Plastic Category"
  )
```

```{r}
glucose_2014_stats <- glucose_2014[glucose_2014$year == 2014, ]
glucose_2014_stats$log_plastics <- log(glucose_2014_stats$plastic + 1)

# Create empty list to store results
results_list <- list()

df_sub <- glucose_2014_stats %>%
  select(log_plastics, mass, glucose)


# Extract relevant data
  df_sub <- glucose_2014_stats[, c(log_plastics, mass, glucose)]
  names(df_sub) <- c("plastic_mass", "mass", "response")
  df_sub <- na.omit(df_sub)
  
  # ----------------------------------------
  # BoxCox model with plastic_mass
  df_sub$plastics <- (df_sub$plastic_mass)
  aov_base <- lm(response ~ plastics, data = df_sub)
  
  boxcox_out <- boxcox(aov_base, lambda = seq(-5, 5, 1/1000), plotit = FALSE)
  lambda <- boxcox_out$x[which.max(boxcox_out$y)]
  
  # Apply transformation
  if (lambda == 0) {
    response_trans <- log(df_sub$response)
  } else {
    response_trans <- df_sub$response^lambda
  }
  
  aov_base_trans <- lm(response_trans ~ plastics, data = df_sub)
  print(tab_model(aov_base_trans))
  print(autoplot(aov_base_trans))
  
  # Extract p-value for plastics
  aov_res <- anova(aov_base_trans)
  p_val <- aov_res["plastics", "Pr(>F)"]
  results_list[[response_name]] <- p_val
  
  # ----------------------------------------
  # Linear model: response ~ mass (body weight)
  lm_mass <- lm(response ~ mass, data = df_sub)
  df_sub$residuals_mass <- resid(lm_mass)
  
  # Plot residuals vs unlogged plastic mass
  p <- ggplot(df_sub, aes(x = plastics, y = residuals_mass)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "blue") +
    labs(
      title = paste("Residuals of", response_name, "vs Plastic Mass"),
      x = "Plastic Mass (g)",
      y = "Residuals from response ~ Body Mass"
    ) +
    theme_classic()
  
  print(p)


  p <- ggplot(df_sub, aes(x = plastics, y = response)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "#0072b2") +
    labs(
      title = paste("Raw data from ", response_name, "vs Plastic Mass"),
      x = "Plastic Mass (g)",
      y = "Response"
    ) +
    theme_classic()
  
  print(p)
  
  file_name <- paste0("plot_", gsub(" ", "_", response_name), ".pdf")

  # Save as PDF
  ggsave(
    filename = file_name,
    plot = p,
    device = "pdf",
    width = 6,
    height = 5
  )

# Convert p-values to dataframe
results_df <- data.frame(
  Response = names(results_list),
  P_Value = unlist(results_list)
)

print(results_df)
```

```{r}
glucose_allyears <- data.frame(
  ID = blood_data$ID,
  glucose = blood_data$glucose,
  plastic = blood_data$Plas.mass,
  mass = blood_data$Mass, 
  year = blood_data$Year)

glucose_2014 <- glucose_allyears[glucose_allyears$year == "2014", ]
glucose_2014$log_plastics <- log(glucose_2014$plastic + 1)

df_sub <- glucose_2014 %>%
  select(log_plastics, mass, glucose)

# Create empty list to store results
results_list <- list()

# Set the response variable name
response_name <- "glucose"

# Subset just the needed columns
df_sub <- glucose_2014_stats %>%
  select(log_plastics, mass, glucose) %>%
  rename(
    plastic_mass = log_plastics,
    response = glucose
  ) %>%
  na.omit()

# ----------------------------------------
# BoxCox model with plastic_mass
df_sub$plastics <- df_sub$plastic_mass
aov_base <- lm(response ~ plastics, data = df_sub)

boxcox_out <- boxcox(aov_base, lambda = seq(-5, 5, 1/1000), plotit = FALSE)
lambda <- boxcox_out$x[which.max(boxcox_out$y)]

# Apply transformation
if (lambda == 0) {
  response_trans <- log(df_sub$response)
} else {
  response_trans <- df_sub$response^lambda
}

aov_base_trans <- lm(response_trans ~ plastics, data = df_sub)
print(tab_model(aov_base_trans))
print(autoplot(aov_base_trans))

# Extract p-value for plastics
aov_res <- anova(aov_base_trans)
p_val <- aov_res["plastics", "Pr(>F)"]
results_list[[response_name]] <- p_val

# ----------------------------------------
# Linear model: response ~ mass (body weight)
lm_mass <- lm(response ~ mass, data = df_sub)
df_sub$residuals_mass <- resid(lm_mass)

# Plot residuals vs unlogged plastic mass
p <- ggplot(df_sub, aes(x = plastics, y = residuals_mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = paste("Residuals of", response_name, "vs Plastic Mass"),
    x = "Plastic Mass (g)",
    y = "Residuals from response ~ Body Mass"
  ) +
  theme_classic()

print(p)

# Raw data plot
p <- ggplot(df_sub, aes(x = plastics, y = response)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "#0072b2") +
  labs(
    title = paste("Raw data from", response_name, "vs Plastic Mass"),
    x = "Plastic Mass (g)",
    y = "Response"
  ) +
  theme_classic()

print(p)

file_name <- paste0("plot_", gsub(" ", "_", response_name), ".pdf")

# Save as PDF
ggsave(
  filename = file_name,
  plot = p,
  device = "pdf",
  width = 6,
  height = 5
)

# Convert p-values to dataframe
results_df <- data.frame(
  Response = names(results_list),
  P_Value = unlist(results_list)
)

results_df

```

```{r}
glucose_2025 <- glucose_allyears[glucose_allyears$year == "2025", ]

# Create empty list to store results
results_list <- list()

# Subset just the needed columns
df_sub <- glucose_2025 %>%
  select(plastic, mass, glucose) %>%
  rename(
    plastic_mass = plastic,
    response = glucose
  ) %>%
  na.omit()

# ----------------------------------------
# BoxCox model with plastic_mass
df_sub$plastics <- df_sub$plastic_mass
aov_base <- lm(response ~ plastics, data = df_sub)

boxcox_out <- boxcox(aov_base, lambda = seq(-5, 5, 1/1000), plotit = FALSE)
lambda <- boxcox_out$x[which.max(boxcox_out$y)]

# Apply transformation
if (lambda == 0) {
  response_trans <- log(df_sub$response)
} else {
  response_trans <- df_sub$response^lambda
}

aov_base_trans <- lm(response_trans ~ plastics, data = df_sub)
print(tab_model(aov_base_trans))
print(autoplot(aov_base_trans))

# Extract p-value for plastics
aov_res <- anova(aov_base_trans)
p_val <- aov_res["plastics", "Pr(>F)"]
results_list[[response_name]] <- p_val

# ----------------------------------------
# Linear model: response ~ mass (body weight)
lm_mass <- lm(response ~ mass, data = df_sub)
df_sub$residuals_mass <- resid(lm_mass)

# Plot residuals vs unlogged plastic mass
p <- ggplot(df_sub, aes(x = plastics, y = residuals_mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = paste("Residuals of", response_name, "vs Plastic Mass"),
    x = "Plastic Mass (g)",
    y = "Residuals from response ~ Body Mass"
  ) +
  theme_classic()

print(p)

# Raw data plot
p <- ggplot(df_sub, aes(x = plastics, y = response)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "#0072b2") +
  labs(
    title = paste("Raw data from", response_name, "vs Plastic Mass"),
    x = "Plastic Mass (g)",
    y = "Response"
  ) +
  theme_classic()

print(p)

file_name <- paste0("plot_", gsub(" ", "_", response_name), ".pdf")

# Save as PDF
ggsave(
  filename = file_name,
  plot = p,
  device = "pdf",
  width = 6,
  height = 5
)

# Convert p-values to dataframe
results_df <- data.frame(
  Response = names(results_list),
  P_Value = unlist(results_list)
)

results_df
```







plastic plots 

```{r}
plastic_mass_plot <- data.frame(
  plastic = blood_data$Plas.mass,
  year = blood_data$Year
)

plastic_mass_plot <- plastic_mass_plot %>%
  filter(year %in% c(2014, 2025))

colours <- viridis(2, begin = 0.50, end = 0.75, alpha = 0.5)
colours <- c("#009e73", "#0072b2")
plastic_mass_plot$year <- factor(plastic_mass_plot$year, levels = c("2025", "2014"))

ggplot(plastic_mass_plot, aes(year, plastic, fill = year, col = year)) +
  geom_flat_violin(position = position_nudge(x = 0.1, y = 0), alpha = 0.8, width = 0.75) +
  geom_point(position = position_jitterdodge(dodge.width = 0.15, jitter.width = 0.15), size = 4, alpha = 0.8) +
  geom_boxplot(color = "black", width = 0.08, outlier.shape = NA, alpha = 0.5, position = position_nudge(x = -0.2, y = 0)) +
  coord_flip() +
  labs(x = "", y = "Relative levels") +
  scale_y_continuous(trans = "sqrt") +
  theme_classic() +
  theme(text = element_text(size = 14)) + 
  scale_fill_manual(values = colours) +
  scale_colour_manual(values = colours) +
  guides(
    colour = guide_legend(order =2),
    shape = "none"
    )
  









rain_cloud_plot <- ggplot(plastic_mass_plot, aes(plastic, year)) 

# Adding a flat violin plot with some horizontal position adjustment and transparency
  geom_flat_violin(position = position_nudge(x = 0.1, y = 0), alpha = 0.8, width = 0.5) +

# Adding points to the plot with jittering and dodging for better visualization, along with some transparency
  geom_point(position = position_jitterdodge(dodge.width = 0.15, jitter.width = 0.15), size = 2, alpha = 0.6) +

# Adding a box plot with black color, no outliers, some transparency, and horizontal position adjustment
  geom_boxplot(color = "black", width = 0.1, outlier.shape = NA, alpha = 0.5, position = position_nudge(x = -0.2, y = 0)) +

# Flipping the x and y axes (transposing) to create a horizontal plot
  coord_flip() +

# Setting the x and y axis labels and adjusting the y-axis scale to a square root transformation
  labs(x = "", y = "Relative levels") +
  scale_y_continuous(trans = "sqrt") +

# Applying a classic theme to the plot
  theme_classic() +

# Removing the legend from the plot
  theme(legend.position = "none") +

# Adjusting the font size for the text elements in the plot
  theme(text = element_text(size = 14)) +

# Setting the color palette for the fill aesthetic
  scale_fill_manual(values = colours) +

# Setting the color palette for the color aesthetic
  scale_colour_manual(values = colours)+
# Relabel x axis
  scale_x_discrete(labels=c("Disease + Drug", "Disease","Drug", "Control"))

# Displaying the plot
rain_cloud_plot






ggplot(plastic_mass_plot, aes(year, plastic, fill = year)) +
  # Flat violin
  geom_flat_violin(position = position_nudge(x = 0.1, y = 0), alpha = 0.8, width = 0.75, colour = NA) +
  # Points with outline matching fill
  geom_point(aes(fill = year, colour = year), 
             shape = 21, 
             position = position_jitterdodge(dodge.width = 0.15, jitter.width = 0.15),
             size = 4, 
             alpha = 0.8) +
  # Boxplot
  geom_boxplot(color = "black", width = 0.08, outlier.shape = NA, alpha = 0.5, position = position_nudge(x = -0.2, y = 0)) +
  coord_flip() +
  labs(x = "", y = "Relative levels") +
  scale_y_continuous(trans = "sqrt") +
  theme_classic() +
  theme(text = element_text(size = 14)) + 
  scale_fill_manual(values = colours) +
  scale_colour_manual(values = colours) +
  guides(
    colour = guide_legend(order = 2),
    shape = "none"
  )


```


```{r}
blood_chem <- blood_data
blood_chem$log_plastic_mass <- log(blood_chem$Plas.mass+1)
years<-c("2014", "2025")
blood_chem_trim <- blood_chem[blood_chem$Year %in% years,]

# safe z-score function (handles NA and zero SD)
zw <- function(x) {
  s <- sd(x, na.rm = TRUE)
  if (is.na(s) || s == 0) return(ifelse(is.na(x), NA_real_, 0))
  (x - mean(x, na.rm = TRUE)) / s
}



blood_chem_trim$glucose_z <- with(blood_chem_trim, ave(glucose, Year, FUN = zw))


lm<-lm(glucose_z ~ Year+log_plastic_mass, data = blood_chem_trim)
autoplot(lm)
tab_model(lm)

blood_chem_trim$mass_group <- ifelse(blood_chem_trim$Mass < 450, "< 450", "≥ 450")

# Plot
ggplot(blood_chem_trim, aes(x = log_plastic_mass, y = glucose_z, colour = factor(Year))) +
  #geom_text(aes(label = ID), vjust = -0.8, size = 3, colour = "black") +
  geom_point(size = 3, shape = 16, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 1, fill = "grey70") +
  labs(
    x = "Plastic Mass (log scale)",
    y = "Glucose Z-score",
    colour = "Year",
    title = "Linear Regression: Glucose Z vs Plastic Mass"
  ) +
  facet_wrap(~ mass_group) +
  scale_colour_manual(values = c("2014" = "#0072b2", "2025" = "#009e73")) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    clip = "off"
  )

```

```{r}
blood_chem_trim$Cal14_Z <- scale(blood_chem_trim$Ca_I)
blood_chem_trim$Cal25_Z <- scale(blood_chem_trim$Ca_Z)

blood_chem_trim$Calcium_z <- rowMeans(blood_chem_trim[, c("Cal14_Z", "Cal25_Z")], na.rm = TRUE)

ggplot(blood_chem_trim, aes(x = log_plastic_mass, y = Calcium_z, colour = factor(Year))) +
  geom_point(size = 3, shape = 16, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 1, fill = "grey70") +
  geom_text_repel(aes(label = ID), size = 3, colour = "black") + 
  labs(
    x = "Plastic Mass (log scale)",
    y = "Calcium Z-score",
    colour = "Year",
    title = "Linear Regression: Calcium Z vs Plastic Mass"
  ) +
  facet_wrap(~ mass_group) +
  scale_colour_manual(values = c("2014" = "#0072b2", "2025" = "#009e73")) +
  theme_classic(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    clip = "off"
  )


```




```{r}

uric_acid <- read.csv("250815 2014 uric acid.csv")



p <- ggplot(uric_acid, aes(x = Plastic, y = Uric)) +
    geom_point() +
    geom_smooth(method = "lm", se = TRUE, color = "#0072b2") +
    labs(
      title = paste("Raw data from ", "Uric acid", "vs Plastic Mass"),
      x = "Plastic Mass (g)",
      y = "Response"
    ) +
    theme_classic()
  
  print(p)


```

